Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-11-08 paper/ch_potentials.tex
update with improvements from diss

\paragraph{Correctness}
Due to~\cite{gssv-erlrn-12}, an up-down path of shortest distance must exist in $G^+$.
Further, $G^+$ can be decomposed into two directed acyclic graphs (DAG) $\gchu$ and $\gchd$ and the up path can be found in $\gchu$ and the down path in $\gchd$.
Observe that the second part of Lazy RPHAST is, in fact, a recursive depth-first search (DFS) on $\gchu$, which relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
Since DFS can be used to find shortest paths on DAGs, the search will find shortest up paths on $\gchu$.
Concatenated with the down paths obtained by the backward search, this yields optimal shortest distances.

@@

We obtain the procedure described Algorithm~\ref{algo:cch_pot}, which utilizes the elimination tree.
For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
Otherwise, the algorithm follows the elimination tree upward until a node with a finalized distance is encountered.
The elimination tree is represented as parent pointer array $P$.
The visited nodes are pushed onto a stack $S$.
This enables the algorithm to enumerate the vertices in reversed order by popping them from the stack.
While popping the nodes, all outgoing upward edges are relaxed.
This finalizes the shortest distances for all vertices on the elimination tree path, including the desired query vertex $v$.

@@

We can also skip some degree-three vertices.
Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
If $|N(w)| > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq v_k \neq a_1$.
Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
If it reaches $b$, $b$'s weight in the queue is set to $D[b]+h_t(b)$.
Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
If neither $y$ nor $b$ are not reached, no queue operation is performed.
Using this method, we avoid pushing every other degree-three vertex into the queue.

@@

In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
We compute the core using Tarjan's algorithm \cite{t-dfslg2-72}.
For every vertex $v$ in the input graph $G$, we store an \emph{attachment vertex} $a_v$ to the core.
For vertices in the core, $a_v=v$.
For every vertex $v$ outside of the core, the attachment vertex $a_v$ is the cut vertex in the core that separates $v$'s component from the core.

The query phase is divided into two steps.
In the first step, we run A* on the subgraph induced by the core and $s$'s component.
Formulated differently, we only consider nodes which are in the core or have the same attachment vertex as $s$.
This can be achieved implicitly by not following edges to nodes without this property.
If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
Otherwise, we find $a_t$.
In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
The final result is the concatenation of both paths.

@@

We observe that the number of queue pushes roughly correlates with running time.
As (C)CH-Potentials and Oracle-A* have the same heuristic values, the number of queue pushes are by construction equal.
Each optimization reduces both queue pushes and running times.
All optimizations yield a combined speedup of around 3.
% Skipping degree three vertices has the smallest impact.
(C)CH-Potentials outperforms ALT by a factor of between six and seven and settle correspondingly fewer vertices.
This is not surprising since ALT computes worse distance estimates.
In contrast, CH-Potentials already computes exact distances with respect to $w_\ell$.
The number of popped vertices is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
This overhead leads to a slowdown of around 1.6.
Thus, CH-Potentials is already very close to the best possible heuristic in this model.
% Very interesting is the comparison with Oracle-A*.
% Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
No competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
CCH-Potentials is slightly slower than CH-Potentials because they use a weight-independent vertex importance order.
% This enables CCH-Potentials to support a customization phase, whereas CH-Potentials cannot.

>>>>

\paragraph{Correctness}
Due to~\cite{gssv-erlrn-12}, an up-down path of shortest distance must exist in $G^+$.
Further, $G^+$ can be decomposed into two directed acyclic graphs (DAG) $\gchu$ and $\gchd$ and the up path can be found in $\gchu$ and the down path in $\gchd$.
The Lazy RPHAST selection finds the shortest down path in $\gchd$.
Now observe that the \textsf{ComputeAndMemoizeDist} function is, in fact, a recursive depth-first search (DFS) on $\gchu$.
Edges $uv$ are relaxed once all upward neighbors of $u$ have been processed, i.e.\ in DFS-post order.
A DFS-post order also is a reverse topological order for $\gchu$.
Therefore, the algorithm relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
Since $\gchu$ is a DAG, this yields shortest up distances in $\gchu$.
Concatenated with the down paths obtained by the backward search, this yields optimal shortest distances.

@@

We obtain the procedure described Algorithm~\ref{algo:cch_pot}, which utilizes the elimination tree.
For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
Otherwise, the algorithm follows the elimination tree upward until a vertex with a finalized distance is encountered.
The elimination tree is represented as parent pointer array $P$.
The visited vertices are pushed onto a stack $S$.
This enables the algorithm to enumerate the vertices in reversed order by popping them from the stack.
While popping the vertices, all outgoing upward edges are relaxed.
This finalizes the shortest distances for all vertices on the elimination tree path, including the desired query vertex $v$.

@@

We can also skip some degree-three vertices.
Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
If $|N(w)| > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq v_k \neq a_1$.
Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
If it reaches $b$, $b$'s weight in the queue is set to $D[b]+h_t(b)$.
Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
If neither $y$ nor $b$ are reached, no queue operation is performed.
Using this method, we avoid pushing every other degree-three vertex into the queue.

@@

In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
We compute the core using Tarjan's algorithm \cite{t-dfslg2-72}.
For every vertex $v$ in the input graph $G$, we store an \emph{attachment vertex} $a_v$ to the core.
For vertices in the core, $a_v=v$.
For every vertex $v$ outside of the core, the attachment vertex $a_v$ is the cut vertex in the core that separates $v$'s component from the core (or a sentinel value $\bot$ for vertices in components not connected to the core).

The query phase is divided into two steps.
First, we run A* on the subgraph induced by the core and $s$'s component.
Formulated differently, we only consider vertices which are in the core or have the same attachment vertex as $s$.
We achieve this implicitly by not following edges to vertices without this property.
If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
Otherwise, we find $a_t$.
In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
The final result is the concatenation of both paths.
When $t$ is not connected to the core ($a_t = \bot$) but $s$ is ($a_s \neq \bot$), we immediately return a distance of $\infty$.

@@

We observe that the number of queue pushes roughly correlates with running time.
Each optimization reduces both queue pushes and running times.
All optimizations yield a combined speedup of around 3.
% Skipping degree three vertices has the smallest impact.
(C)CH-Potentials outperforms ALT by a factor of between six and seven and settle correspondingly fewer vertices.
This is not surprising since ALT computes worse distance estimates.
In contrast, CH-Potentials already computes exact distances with respect to $w_\ell$.
As (C)CH-Potentials and Oracle-A* have the same heuristic values, the number of queue pushes are by construction equal.
The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
This overhead leads to a slowdown of around 1.6.
Thus, CH-Potentials is already very close to the best possible heuristic in this model.
% Very interesting is the comparison with Oracle-A*.
% Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
No competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
CCH-Potentials is slightly slower than CH-Potentials because they use a weight-independent vertex importance order.
% This enables CCH-Potentials to support a customization phase, whereas CH-Potentials cannot.
